---
title: "Document Title"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

Markdown files can be used to directly generate webpages, PDFs and presentations slides. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

With R Markdown, you can include content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

> However, when producing statistical reports it is rare that we want to embed code chunks, therefore below you can see how we can include tables and visualisations of data.

# Including data

## Tables

You can include tables, for example:

```{r, echo=FALSE}
#Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
knitr::kable(mtcars[1:3,1:3])
```

## Plots

You can also show visualisations and plots:

```{r pressure, echo=FALSE}
#Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
plot(pressure)
```

### Details
You can also add additional elements like listing key points from the data:

* First bullet
  * Nested bullets are also available
* Second bullet
* Third bullet

To generate into a document, you need to select 'knit'.