---
title: "Editing and Imputation"
output: html_document
---

## Editing and Imputation - code examples

The following code relates to the **Methods of reviewing data** section of the
**Editing and Imputation** learning resource on the [Analysis for Action]([PLACEHOLDER])
platform.


## Troubleshooting

If you encounter any issues running the code snippets, consider the following
troubleshooting steps:
- Ensure you have followed the setup instructions in the
[README.md](../../../README.md) file including `install.packages()` for
required packages.
- Run code from top to bottom to ensure all dependencies are met.
- Restart your R session and clear all outputs.
- For more detailed troubleshooting guidance, refer to the
[troubleshooting document]([PLACEHOLDER]).

## Setup

Before running the code snippets, ensure you have:
- followed the setup instructions in the [README.md](../../../README.md) file,
- **[add further instructions as needed]**,
- run the code block below.

```{r setup, message=FALSE}
# Load in the following packages

library(tidyverse)
library(editrules)
library(deducorrect)
library(VIM)
library(mice)
library(lattice)
library(lme4)
library(igraph)

# Use here::here() to construct paths, do not setwd()

# Load in data
# assign to the name survey_data
# TO BE CHANGED WHEN CODE FINALISED
survey_data <- read.csv("demo_data.csv")
```

## Understanding the data

First take a look at the structure of the data set and the first 10 observations.

```{r}
# Use dplyr::glimpse() function to look at the structure of the data set

dplyr::glimpse(survey_data)


# Look at first 10 observations of survey_data data set 

head(survey_data, n = 10)
```

## Reviewing and editing the data 

Reviewing data and editing it aims to improve the quality and accuracy of data. Here, the following aspects of editing data are covered:

- Data entry
- Automatic editing
- Error localization

### Data entry

This stage involves quality assurance of micro data to identify and resolve obvious, known issues such as duplicate records, incorrect formatting, identifying missing values and looking for invalid values.

```{r}
# Look at all the unique values by each variable 

sapply(survey_data, function(x) unique(x))

# As you can see, there are some invalid values in the data set. For example, the variable "region" has a value of "Unknown" and the variable household_size has a value of "-". 
# In R, only NA is represented as a missing value, so these values need to be addressed. 
# Assign all missing values to NA

survey_data[survey_data == c("Unknown","-")] <- NA


# Understand the missing value pattern

mice::md.pattern(survey_data, plot = FALSE)


# Plot the missing values
# col is a vector giving the colours to be used for observed and missing
# numbers give the proportion or frequencies of the different combinations 
# sortVars variables should be sorted by the number of missing values
# labels on the x-axis
# cex.axis is the character expansion factor to be used for the axis annotation
# gap is a numeric value giving the distance between the two plots in margin lines
# ylab are the labels on the y-axis


VIM::aggr(survey_data, col = mice::mdc(1:2),  numbers = TRUE, sortVars = TRUE, 
          labels=names(data), cex.axis = 0.7, gap = 1, oma = c(8,5,3,3),
          ylab=c("Proportion of missingness","Missingness Pattern"))

```

The pink colour shows the missing values while the blue shows the present values. The plot helps us understanding that almost 46% of the samples are not missing any information, 24% of samples are missing the hospital_admission value while other values are present, 0.12% of samples are missing the hospital_admission, age and date_of_birth values while other values are present, and the remaining ones show other missing patterns.

Notice: If you are getting a plot which does not fit the screen, then please change the cex.axis, gap values and use the function oma() to fix outer margin area according to your screen e.g. cex.axis = 0.6, gap = 1 and oma = c(10,5,3,3).

You can also find all rows which have missing values as shown below:

```{r}
# complete case returns a logical vector indicating which cases are complete, i.e., have no missing values
# instead, use ! which acts as a negator so it will return a logical vector where cases are not complete
# select all columns by leaving space after the comma

missing_values <- survey_data[!complete.cases(survey_data), ]


# First six rows with missing value

head(x = missing_values)
```

#### Finding duplicates

```{r}
# Show the repeat entries

survey_data[duplicated(survey_data),]

# Keep only distinct data entries

survey_data <- distinct(survey_data)
```

#### Date formatting

```{r}
# Correcting date formatting
# Convert date from character variable to date format

survey_data$date_of_birth <- as.Date(survey_data$date_of_birth, format = "%d/%m/%Y")

# Additionally, can split date of birth into day, month, year columns 

survey_data <- survey_data %>% 
  mutate(dob_year = as.integer(format(survey_data$date_of_birth, "%Y")),
         dob_month = as.integer(format(survey_data$date_of_birth, "%m")),
         dob_day = as.integer(format(survey_data$date_of_birth, "%d")))
```



## Automatic editing

In a real-life data set, there are simple errors that need to be treated automatically. 
For example, a persons age cannot be negative and an under-aged person cannot have a drivers license. Such knowledge can be expressed as rules or constraints. 
In data editing literature these rules are referred to as edit rules or edits, in short. The editrules package can be used to define rules on categorical, numerical or mixed-type data sets which each record must obey. 
Automatic editing can be an easy way to improve data quality, but you must be confident that your assumptions are correct, otherwise your automatic corrections could introduce more error.

Below shows an example of how some restrictions can be applied to the age column in survey_data using the editrules package.

```{r}
# Age --------

# Set the rule where age should be over 18 and less than and equal to 80 only

edit_rules <- editrules::editset(c("age >=0", "age <= 80"))

# Display the results

edit_rules

# Check if the rules are violated 

violate_edits <- editrules::violatedEdits(edit_rules, dat = survey_data)

# Display first six rows

violate_edits[1:6,]
```

The first column shows the records (rows), the second column num1 (age >=0) shows that all values are FALSE meaning that the rule of age being positive has not been violated (age column has positive values for the first 6 records). 
The second column num2 (age <= 80) shows that the 2nd record violates the rule because the logical value is TRUE suggesting that the age of the 2nd person is greater than 80. Indeed, they are 81.


In the next example, some restrictions are set on the vaccination status column. 

```{r}

# Vaccination status --------

# Set the rule where vaccination status should be non-negative and less than and equal to 3

edit_rules <- editrules::editset(c("vaccination_status >=0", "vaccination_status <= 3"))

# Check if the rules are violated 
# the function returns a matrix

edits_violated <- editrules::violatedEdits(edit_rules, dat = survey_data)

# Display the first 6 rows

edits_violated[1:6,]

# Convert the variable to a data frame

violated_data <- data.frame(edits_violated)

# Filter the rows where either of the edit rules have been violated
# once you create a data frame you can see the column names by using $ sign

violated_data %>% 
  dplyr::filter(num1 == TRUE | num2 == TRUE)

```

The second column num2 (vaccination_status <= 3) shows that the 5th and 821st record violates the rule because the logical value is TRUE suggesting that the vaccination_status of both people is greater than 3.
This can be double checked.

```{r}
# Filter the 5th and 821st person in the data set

survey_data[c(5,821),]
```

As you can see the value is 10 (instead of 1) for the 5th person and 20 instead of 2 for the 821st person. 
Here, the assumption will be made that the person made a typing error where the value should be 1 and 2 respectively. 
The package deducorrect can be used here, as this package fixes inconsistent observations by altering invalid values in the record based on information from valid values.

```{r}
# Specify the rule and the correction 

correction_rules <- deducorrect::correctionRules(expression(if(vaccination_status > 16) {
      vaccination_status <- vaccination_status/10
}))

# Apply simple replacement rules to a data.frame

correction <- deducorrect::correctWithRules(rules = correction_rules, dat = survey_data)

# list of alterations

correction$corrections
```

# Error localisation: Editing in large scale datasets

NOTE FOR TESTER: This code is not fully finalised yet, so you may run into errors if you run it. The code is still being worked on and will be finalised soon.

The above code shows how to set restrictions and correct those rows which violate the restrictions.

However in the real world, large data sets are used, which this means there are more variables, and more edit rules.
Rather than having one specific condition, all conditions for all data types i.e. numeric, categorical and mixed data) can be saved in one text file. 

```{r}

# Read the text file containing rules 

edit_rules_list <- editfile("edits.txt")

# Check if the rules are violated

violated_rules <- editrules::violatedEdits(edit_rules_list, survey_data)

# Display the summary

summary(violated_rules)



```