---
title: "Basic Data Visualisation, How to make bar charts in R"
output: html_document
---

This block contains instructions on how to fill in this template.

**Please delete this block when creating your own file.**

This template should be used to create code snippet notebooks for each chapter in a learning resource. Follow the instructions below to complete the template: - Replace content in square brackets (e.g., [unit_name], [link], [chapter name]) with the relevant information. For example the title block below could be changed from "[unit_name], [chapter_name] - Code snippets" to "Basic Data Visualisation, Creating Bar Charts - Code snippets". - In the setup section add any instructions for users to follow before running the code snippets, such as installing packages or downloading data files. Add any necessary import statements or setup code in the first code cell. - Code snippets should be grouped into sections according to the sections in the learning resource. - Each code snippet from the learning resource should be placed in its own code cell under the relevant section, with any necessary context provided in a markdown block above each code cell. - Follow [tidyverse](https://style.tidyverse.org/) style guide for R code formatting. - Whenever you need to reference a file in [learning_resources](http://_vscodecontentref_/1), use:

``` r
file_path <- file.path(learning_resources_dir, "data", "filename.csv")
```

-   Definitions:
    -   "Unit name" refers to the unit name, which is also the name of the folder this python folder is in.
    -   "Chapter name" refers to the specific chapter that the code snippets are in. Chapters are the highest level headings in the learning resource.
    -   "Section name" refers to subheadings within chapters that code snippets are in. ---

## Basic Data Visualisation, How to make bar charts in R - Code examples

The following code relates to the **How to make bar charts in R** section of the **Basic Data Visualisation** learning resource on the [Analysis for Action](%5BPLACEHOLDER%5D) platform.

Use this code to get to grips with the basics of making bar charts in R as per the best practice covered in the written guidance. You are welcome to copy and adapt the code to meet your needs, but make sure that any outputs also align with any style requirements for your organisation.

The code applies each change separately and does not create any outputs.

## Troubleshooting

If you encounter any issues running the code snippets, consider the following troubleshooting steps: - Ensure you have followed the setup instructions in the [README.md](../../../README.md) file including `install.packages()` for required packages. - Run code from top to bottom to ensure all dependencies are met. - Restart your R session and clear all outputs. - For more detailed troubleshooting guidance, refer to the [troubleshooting document](%5BPLACEHOLDER%5D).

## Setup

Before running the code snippets, ensure you have: - followed the setup instructions in the [README.md](../../../README.md) file, - **[add further instructions as needed]**, - run the code block below.

```{r setup, message=FALSE}
# This chunk should include:
# * library() calls for required packages
# * paths to data files
# * any other setup required

library(readr)
library(here)
library(dplyr)
library(tidyr)
library(janitor)
library(ggplot2) # The scripts uses ggplot2 version 4.0.0
library(svglite)
library(scales)
library(plotly)
library(htmlwidgets)
library(forcats)

# Use here::here() to construct paths, do not setwd()
here::here()
vuln <- read.csv(here::here("learning_resources", "data", "vulnerable.csv"))
```

## Introduction to ggplot2

The main package you will use to create a bar chart in this file is ggplot2.
The package works using layers, creating the base chart and then adding layers to change titles, colours and more.

## Prepare your data

This code uses the 'vulnerable' dataset.

The raw data is too large to use in a bar chart. You need to filter it to make a smaller dataframe first.

```{r}
# Prepare your data to make it suitable for a bar chart

vuln_2007_top_6 <- vuln %>%
  # Filter the data to just the year 2007
  filter(year %in% ("2007")) %>%
  # Select just the country and vulnerable_pop columns
  select(country, vulnerable_pop) %>%
  # Select only the rows with the 6 highest vulnerable_pop value
  dplyr::top_n(n = 6)

# To Display the data

vuln_2007_top_6
```

## Make the base bar chart

```{r}
# Creating a gglot2 bar chart
# Used vuln_2007_top_6 dataframe, country displayed on the x-axis vulnerable_pop displayed the on y-axis
vuln_bar <- ggplot(vuln_2007_top_6, aes(x = country, y = vulnerable_pop)) +
  geom_bar(stat = "identity") +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  # label_number used to remove scientific notation
  scale_y_continuous(labels = label_number(scale = 1)) +
  # Set y-axis title to horizontal
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.06,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -70)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5))


vuln_bar
```

## Make the chart horizontal

```{r}
vuln_bar_horizontal <- ggplot(vuln_2007_top_6, aes(x = country, y = vulnerable_pop)) +
  geom_bar(stat = "identity") +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank()) +
  # Flip the chart so that the bars are horizontal
  coord_flip() +
  # label_number used to remove scientific notation
  scale_y_continuous(labels = label_number(scale = 1)) +
  # Set y-axis title to horizontal
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.06,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -70)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 50, r = 5, b = 5))

vuln_bar_horizontal
```

## Rename axes

```{r}
vuln_bar_titles <- ggplot(vuln_2007_top_6, aes(x = country, y = vulnerable_pop)) +
  geom_bar(stat = "identity") +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  # Add axis titles
  labs(x = "Country", y = "Vulnerable Population") +
  # label_number used to remove scientific notation
  scale_y_continuous(labels = label_number(scale = 1)) +
  # Set y-axis title to horizontal
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.06,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -70)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5))
  
vuln_bar_titles
```

## Set a maximum value for the y-axis

```{r}
vuln_bar_y_axis_changed <- ggplot(vuln_2007_top_6, aes(x = country, y = vulnerable_pop)) +
  geom_bar(stat = "identity") +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  # label_number used to remove scientific notation
  scale_y_continuous(labels = label_number(scale = 1)) +
  # Set a limit for the y-axis
  ylim(0, 400000000) +
  # Set y-axis title to horizontal
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.06,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -70)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5))

vuln_bar_y_axis_changed
```

## Reorder bars

```{r}
vuln_2007_top_6_arranged <- vuln_2007_top_6 %>% 
  # Order the data frame rows from largest to smallest
  dplyr::arrange(desc(vulnerable_pop))

# dplyr::arrange(vulnerable_pop) to reverse the order

vuln_2007_top_6_arranged$country <- factor(
  vuln_2007_top_6_arranged$country,
  levels = vuln_2007_top_6_arranged$country
)

# Create the bar chart using the arranged data
vuln_bar_arranged <- ggplot(vuln_2007_top_6_arranged, aes(x = country, y = vulnerable_pop)) +
  geom_bar(stat = "identity") +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  # label_number used to remove scientific notation
  scale_y_continuous(labels = label_number(scale = 1)) +
  # Set y-axis title to horizontal
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.06,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -70)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5))

vuln_bar_arranged
```

## Change bar width

```{r}
vuln_bar_change_width <- ggplot(vuln_2007_top_6, aes(x = country, y = vulnerable_pop)) +
  geom_bar(stat = "identity",
           # Change the bar width
           width = 0.6) +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  # label_number used to remove scientific notation
  scale_y_continuous(labels = label_number(scale = 1)) +
  # Set y-axis title to horizontal
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.06,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -70)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5))

vuln_bar_change_width
```

## Change colours of bars

```{r}
vuln_bar_change_bar_colour <- vuln_bar <- ggplot(vuln_2007_top_6, aes(x = country, y = vulnerable_pop)) +
  # Change all bars to the same colour
  geom_bar(stat = "identity", fill = "#324027") +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  # label_number used to remove scientific notation
  scale_y_continuous(labels = label_number(scale = 1)) +
  # Set y-axis title to horizontal
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.06,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -70)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5))

vuln_bar_change_bar_colour
```

## Change bar colours (focus chart)

```{r}
# Create a new DataFrame which is a copy of vuln_2007_top_6 with a column called highlight which contains the value TRUE if the row contains the maximum vulnerable_pop value, all other rows contain the value FALSE in the highlight column
vuln_focus <- vuln_2007_top_6 %>%
  mutate(highlight = vulnerable_pop == max(vulnerable_pop))

# Create a bar chart with the vuln_focus data 
vuln_bar_focus_chart <- ggplot(vuln_focus, aes(x = country, y = vulnerable_pop, fill = highlight)) +
  geom_bar(stat = "identity") +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  # Change the bar colours based off the whether highlight is TRUE or FALSE
  scale_fill_manual(values = c("TRUE" = "#1e2a16", "FALSE" = "#959595"),
                    guide = "none") +
  # label_number used to remove scientific notation
  scale_y_continuous(labels = label_number(scale = 1)) +
  # Set y-axis title to horizontal
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.06,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -70)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5))

vuln_bar_focus_chart
```

## Add labels to bars

```{r}
vuln_bar_add_labels <- ggplot(vuln_2007_top_6, aes(x = country, y = vulnerable_pop)) +
  geom_bar(stat = "identity") +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  # Add labels to bars containing the vulnerable population of each country
  geom_text(aes(label = vulnerable_pop), 
            # Adjust the position of the label
            vjust = -0.25,
            # Adjust the text size of the label
            size = 4) +
  # label_number used to remove scientific notation
  scale_y_continuous(labels = label_number(scale = 1)) +
  # Set y-axis title to horizontal
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.06,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -70)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5))

vuln_bar_add_labels
```

## Specify the number of axis break

```{r}
vuln_bar_custom_axis_breaks <- ggplot(vuln_2007_top_6, aes(x = country, y = vulnerable_pop)) +
  geom_bar(stat = "identity") +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  # Custom axis breaks are used, y-axis is breaks are every 50000000 people
  scale_y_continuous(breaks = c(0, 50000000, 100000000, 150000000, 200000000, 250000000, 300000000, 350000000),
  # The function expansion is used to increase the vertical height of the plot by 40% to show the extra axis breaks
  expand = expansion(mult = c(0, 0.4)),
  # label_number used to remove scientific notation
  labels = label_number(scale = 1)) +
  # Set y-axis title to horizontal
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.06,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -70)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5))
                              
vuln_bar_custom_axis_breaks 
```

# Add number suffixes (Thousands)

```{r}
vuln_bar_thousands <- ggplot(vuln_2007_top_6, aes(x = country, y = vulnerable_pop)) +
  geom_bar(stat = "identity") +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  # Divide the axis breaks by 1000 and add the suffix k
  scale_y_continuous(labels = label_number(scale = 1e-3, suffix = "k")) +
  # Change the y-axis label
  labs(y = "Vulnerable Population (Thousands)") +
  # Set y-axis title to horizontal
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.06,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -170)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5))

vuln_bar_thousands
```

# Add number suffixes (Millions)

```{r}
vuln_bar_millions <- ggplot(vuln_2007_top_6, aes(x = country, y = vulnerable_pop)) +
  geom_bar(stat = "identity") +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  # Divide the axis breaks by 1000000 and add the suffix M
  scale_y_continuous(labels = label_number(scale = 1e-6, suffix = "M")) +
  # Change the y-axis label
  labs(y = "Vulnerable Population (Millions)") +
  # Set y-axis title to horizontal
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.06,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -150)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5))

vuln_bar_millions
```

# Add number suffixes (Billions)

```{r}
vuln_bar_billions <- ggplot(vuln_2007_top_6, aes(x = country, y = vulnerable_pop)) +
  geom_bar(stat = "identity") +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  # Divide the axis breaks by 1000000000 and add the suffix B
  scale_y_continuous(labels = label_number(scale = 1e-9, suffix = "B")) +
  # Change the y-axis label
  labs(y = "Vulnerable Population (Billions)") +
  # Set y-axis title to horizontal
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.06,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -150)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5))

vuln_bar_billions
```

## Exporting (.svg file type)

```{r}
vuln_bar <- ggplot(vuln_2007_top_6, aes(x = country, y = vulnerable_pop)) +
  geom_bar(stat = "identity") +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  # label_number used to remove scientific notation
  scale_y_continuous(labels = label_number(scale = 1)) +
  # Set y-axis title to horizontal
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.03,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -70)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5))

ggsave(
  # Set the file name
  filename = "vuln_pop_2007.svg",
  # Select the plot to save
  plot = vuln_bar,
  # Set the save path
  path = "D:/Temp",
  # Set aspect ratio of plot, an aspect ratio of 16:9 will fill 1920x1080 displays
  width = 16,
  height = 9
)
```

# Exporting (.html file format) (issues with axis labels)

```{r}
vuln_bar <- ggplot(vuln_2007_top_6, aes(x = country ,y = vulnerable_pop)) +
  geom_bar(stat = "identity") +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  # label_number used to remove scientific notation
  scale_y_continuous(labels = label_number(scale = 1)) +
  # Set y-axis title to horizontal
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.06,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -70)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5))

vuln_bar_html <- ggplotly(vuln_bar)

saveWidget(vuln_bar_html, file = "D:/Temp/vuln_pop_2007.html")
```

## Exercise

Use this exercise to test what you’ve learned about making bar charts. You can adapt much of the code covered in this section, but some basic data manipulation will also be required which has not been covered here.

Create a bar chart in R. Use the vulnerable dataset as your input data.

For an extra challenge, consider setting yourself an hour’s time limit to help simulate an emergency.

The bar chart should:

-   Show the size of vulnerable populations in Argentina, Malawi, Nepal, and the United Kingdom in the year 1997
-   Be formatted so that:
    -   The x-axis label is: Country 
    -   The y-axis label is: Vulnerable Population (Millions) 
    -   The text size of the axis titles is 12
    -   theme_minimal is used to set the background and gridline colours
    -   Vertical gridlines are removed
    -   The largest bar is HEX code #324028, the rest are HEX code #D4DDD6
    -   The width of the bars is 0.6
    -   The bars are ordered from largest to smallest
    -   The y-axis uses millions suffix (for example an axis break with a value of 6000000 would show as 6M)
    -   The chart can be exported as an .svg file or a .html file with the file name:
        vuln_pop_partner_countries 
    -   The chart is saved to the D:/Temp folder
    -   The .svg chart has an aspect ratio of 16:9

When you have finished making your bar chart, compare it against the relevant code and solution in the 'exercise_solutions' subfolder. If you have followed all the guidance correctly, this is approximately how your output should look.

## Extras

Making clustered, stacked and 100% stacked bar charts are is similar to creating regular bar charts, they tend to require either small changes to the data or changes to the ggplot layers. The majority of the formatting options covered earlier in this markdown can be use when creating clustered or stack bar charts

## Clustered Bar Chart

```{r}
# Prepare your data to make it suitable for a bar chart

vuln_2002_and_2007_top_6 <- vuln %>%
  # Filter the data to just the years 2002 and 2007
  filter(year %in% c(2002,2007)) %>%
  
  mutate(year = as.factor(year)) %>%
  # Select just the country, year and vulnerable_pop columns
  select(country, year, vulnerable_pop) %>%
  # Group the data by year
  group_by(year) %>%
  # Select only the rows with the 6 highest vulnerable_pop value for each year
  slice_max(order_by = vulnerable_pop, n = 6) %>%

  ungroup()

# To Display the data

vuln_2002_and_2007_top_6
```

```{r}
vuln_cluster <- ggplot(vuln_2002_and_2007_top_6, aes(x = country, y = vulnerable_pop, fill = year)) +
  # position = "dodge" sets the chart to a clustered bar chart
  geom_bar(stat = "identity", position = "dodge") +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  # Change the colours of bars
  scale_fill_manual(
    values = c("2002" = "#48553F", "2007" = "#F59342")
  ) +
  # label_number used to remove scientific notation
  scale_y_continuous(labels = label_number(scale = 1)) +
  # Set y-axis title to horizontal
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.06,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -70)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5))

vuln_cluster
```

## Stacked bar chart

```{r}
# Prepare your data to make it suitable for a stacked bar chart

vuln_all_years_top_6 <- vuln %>%
  filter(country %in% c("Pakistan","Brazil","Indonesia","United States","India","China")) %>%
  
  select(country, year, vulnerable_pop)

vuln_all_years_top_6_ordered <- vuln_all_years_top_6 %>%
  mutate(
    country = fct_reorder(
      country,
      if_else(year == 2007, vulnerable_pop, NA_real_),
      na.rm = TRUE),
      year = as.numeric(as.character(year))
  )

vuln_all_years_top_6_ordered
```

```{r}
vuln_stacked_bar <- ggplot(vuln_all_years_top_6_ordered, aes(x = year, y = vulnerable_pop, fill = country)) +
  geom_bar(stat = "identity") +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  # Change the colours of bars
  scale_fill_manual(
    values = c("Pakistan" = "#1E2A1F", "Brazil" = "#D17E38", "Indonesia" = "#444443", "United States" = "#3EA27F", "India" = "#752111", "China" = "#618DD3")
  ) +
  scale_x_continuous(breaks = vuln_all_years_top_6_ordered$year) +
  # label_number used to remove scientific notation
  scale_y_continuous(labels = label_number(scale = 1)) +
  # Set y-axis title to horizontal
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.06,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -70)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5))

vuln_stacked_bar
```

## 100% Stacked bar chart

```{r}
# Prepare your data to make it suitable for a 100% stacked bar chart

vuln_1997_2002_2007_percent <- vuln %>%
  # Filter the data to just the years 1997, 2002 and 2007
  filter(year %in% c(1997,2002,2007)) %>%
# Countries selected to show a more clear difference with bars, with the previous countries all bars were very similar  
  filter(country %in% c("Afghanistan", "Canada", "Iraq", "Peru", "Romania")) %>%
  # Group the data by year
  group_by(year) %>%
  # Create a percent column which contains each countrys vulnerable population as a percentage of a total vulnerable population of the 5 countries for that year
  # Create a vuln_totals column which contains the total vulnerable population of the 5 countries for each year
  mutate(
    percent = vulnerable_pop / sum(vulnerable_pop),
    vuln_totals = sum(vulnerable_pop)) %>%

  ungroup() %>%
  
  mutate(year = as.factor(year)) %>%
  # Select just the country, year, vulnerable_pop, percent and vuln_totals columns
  select(country, year, vulnerable_pop, percent, vuln_totals)

# Create seperate table for year totals to make labels with

year_totals <- vuln_1997_2002_2007_percent %>%
  
  distinct(year, vuln_totals)

# To Display the data

vuln_1997_2002_2007_percent

year_totals
```

```{r fig.width=16, fig.height=6}
vuln_stacked_bar_percent <- ggplot(vuln_1997_2002_2007_percent, aes(x = year, y = vulnerable_pop, fill = country)) +
  geom_col(position = "fill") +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove horizontal gridlines
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank()) +
  # Add the percentages of the total population inside the bars
  geom_text(
    aes(label = percent(percent, accuracy = 0.01)),
    position = position_fill(vjust = 0.5),
    color = "#FFFFFF",
    size = 4
  ) +
  # Create labels at the end of bars showing the total population for each year
  geom_text(
    data = year_totals,
    aes(x = year, y = 1.04, label = comma(vuln_totals)),
    inherit.aes = FALSE,
    size = 4,
    fontface = "bold"
  ) +
  coord_flip() +
  # Change the colours of bars
  scale_fill_manual(
    values = c("Afghanistan" = "#1E2A16",
               "Canada" = "#D17E38",
               "Iraq" = "#444443",
               "Peru" = "#3EA27F",
               "Romania" = "#752111")
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  # Create labels above the populations
  annotate(
  "text",
  x = factor("2007", levels = levels(vuln_1997_2002_2007_percent$year)),
  y = 1.08,
  label = "Total vulnerable population 2007:",
  fontface = "bold",
  hjust = 0.6,
  vjust = -2,
  size = 2.2) +
  annotate(
  "text",
  x = factor("2002", levels = levels(vuln_1997_2002_2007_percent$year)),
  y = 1.08,
  label = "Total vulnerable population 2002:",
  fontface = "bold",
  hjust = 0.6,
  vjust = -2,
  size = 2.2) +
    annotate(
  "text",
  x = factor("1997", levels = levels(vuln_1997_2002_2007_percent$year)),
  y = 1.08,
  label = "Total vulnerable population 1997:",
  fontface = "bold",
  hjust = 0.6,
  vjust = -2,
  size = 2.2) +
  # Set y-axis title to horizontal
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.03,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -20)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5))
  
vuln_stacked_bar_percent
```

