---
title: "Basic Data Visualisation, How to make line charts in R"
output: html_document
---

This block contains instructions on how to fill in this template.

**Please delete this block when creating your own file.**

This template should be used to create code snippet notebooks for each chapter in a learning resource. Follow the instructions below to complete the template: - Replace content in square brackets (e.g., [unit_name], [link], [chapter name]) with the relevant information. For example the title block below could be changed from "[unit_name], [chapter_name] - Code snippets" to "Basic Data Visualisation, Creating line charts - Code snippets". - In the setup section add any instructions for users to follow before running the code snippets, such as installing packages or downloading data files. Add any necessary import statements or setup code in the first code cell. - Code snippets should be grouped into sections according to the sections in the learning resource. - Each code snippet from the learning resource should be placed in its own code cell under the relevant section, with any necessary context provided in a markdown block above each code cell. - Follow [tidyverse](https://style.tidyverse.org/) style guide for R code formatting. - Whenever you need to reference a file in [learning_resources](http://_vscodecontentref_/1), use:

``` r
file_path <- file.path(learning_resources_dir, "data", "filename.csv")
```

-   Definitions:
    -   "Unit name" refers to the unit name, which is also the name of the folder this python folder is in.
    -   "Chapter name" refers to the specific chapter that the code snippets are in. Chapters are the highest level headings in the learning resource.
    -   "Section name" refers to subheadings within chapters that code snippets are in. ---

## Basic Data Visualisation, How to make line charts in R - Code examples

The following code relates to the **How to make line charts in R** section of the **Basic Data Visualisation** learning resource on the [Analysis for Action](%5BPLACEHOLDER%5D) platform.

Use this code to get to grips with the basics of making line charts in R as per the best practice covered in the written guidance. You are welcome to copy and adapt the code to meet your needs, but make sure that any outputs also align with any style requirements for your organisation.

The code applies each change separately and does not create any outputs.

## Troubleshooting

If you encounter any issues running the code snippets, consider the following troubleshooting steps: - Ensure you have followed the setup instructions in the [README.md](../../../README.md) file including `install.packages()` for required packages. - Run code from top to bottom to ensure all dependencies are met. - Restart your R session and clear all outputs. - For more detailed troubleshooting guidance, refer to the [troubleshooting document](%5BPLACEHOLDER%5D).

## Setup

Before running the code snippets, ensure you have: - followed the setup instructions in the [README.md](../../../README.md) file, - **[add further instructions as needed]**, - run the code block below.

```{r setup, message=FALSE}
# This chunk should include:
# * library() calls for required packages
# * paths to data files
# * any other setup required
library(readr)
library(here)
library(dplyr)
library(tidyr)
library(janitor)
library(ggplot2) # The scripts uses ggplot2 version 4.0.0
library(scales)
library(svglite)
library(plotly)
library(htmlwidgets)
library(ggrepel)

# Use here::here() to construct paths, do not setwd()
here::here()
vuln <- read.csv(here::here("learning_resources", "data", "vulnerable.csv"))
```

## Introduction to ggplot2

The main package you will use to create a line chart in this file is ggplot2.
The package works using layers, creating the base chart and then adding layers to change titles, colours and more.

## Prepare your data

This code uses the 'vulnerable' dataset.

The raw data is too large to use in a line chart. You need to filter it to make a smaller dataframe first.

```{r}
# Prepare your data to make it suitable for a line chart

vuln_4_country <- vuln %>%
  
  filter(country %in% c("Argentina", "Malawi", "Nepal","United Kingdom")) %>%
  
  select(year, vulnerable_pop, country)

# To Display the data

vuln_4_country

# Prepare data used to create labels

labels_df <- vuln_4_country %>%
  
  group_by(country) %>%
  
  slice_max(year)

labels_df
```

## Make the base line chart

```{r}
vuln_line <- ggplot(vuln_4_country, aes(x = year, y = vulnerable_pop, colour = country)) +
  geom_line() +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  scale_y_continuous(labels = label_number(scale = 1)) +
  # Change the colours of lines to differentiate between countries
  scale_colour_manual(values = c("United Kingdom" = "#1E2A16",
                                 "Argentina" = "#D17E38",
                                 "Nepal" = "#444443",
                                 "Malawi" = "#3EA27F"),
  ) +
  # Remove legend set y-axis title to horizontal
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.06,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -70)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5)) +
  theme(legend.position = "none") +
  # Add labels to lines
  geom_text_repel(data = labels_df,
                  aes(label = country),
                  nudge_y = 500000,
                  colour = "#000000")

vuln_line
```

## Add axis titles

```{r}
vuln_line_titles <- ggplot(vuln_4_country, aes(x = year, y = vulnerable_pop, colour = country)) +
  geom_line() +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  scale_y_continuous(labels = label_number(scale = 1)) +
  labs(x = "Year", y = "Vulnerable Population") +
  # Change the colours of lines to differentiate between countries
  scale_colour_manual(values = c("United Kingdom" = "#1E2A16",
                                 "Argentina" = "#D17E38",
                                 "Nepal" = "#444443",
                                 "Malawi" = "#3EA27F"),
  ) +
  # Remove legend set y-axis title to horizontal
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.06,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -100)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5)) +
  theme(legend.position = "none") +
  # Add labels to lines
  geom_text_repel(data = labels_df,
                  aes(label = country),
                  nudge_y = 500000,
                  colour = "#000000")

vuln_line_titles
```

## Change line thickness

```{r}
vuln_line <- ggplot(vuln_4_country, aes(x = year, y = vulnerable_pop, colour = country)) +
  geom_line(size = 1) +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  scale_y_continuous(labels = label_number(scale = 1)) +
  labs(x = "Year", y = "Vulnerable Population") +
  # Change the colours of lines to differentiate between countries
  scale_colour_manual(values = c("United Kingdom" = "#1E2A16",
                                 "Argentina" = "#D17E38",
                                 "Nepal" = "#444443",
                                 "Malawi" = "#3EA27F"),
  ) +
  # Remove legend set y-axis title to horizontal
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.06,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -100)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5)) +
  theme(legend.position = "none") +
  # Add labels to lines
  geom_text_repel(data = labels_df,
                  aes(label = country),
                  nudge_y = 500000,
                  colour = "#000000")

vuln_line
```

## Add number suffixes (Thousands)

```{r}
vuln_line_thousands <- ggplot(vuln_4_country, aes(x = year, y = vulnerable_pop, colour = country)) +
  geom_line() +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  labs(x = "Year", y = "Vulnerable Population (Thousands)") +
  scale_y_continuous(labels = label_number(scale = 1e-3, suffix = "k")) +
  # Change the colours of lines to differentiate between countries
  scale_colour_manual(values = c("United Kingdom" = "#1E2A16",
                                 "Argentina" = "#D17E38",
                                 "Nepal" = "#444443",
                                 "Malawi" = "#3EA27F"),
  ) +
  # Remove legend set y-axis title to horizontal
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.06,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -170)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5)) +
  theme(legend.position = "none") +
  # Add labels to lines
  geom_text_repel(data = labels_df,
                  aes(label = country),
                  nudge_y = 500000,
                  colour = "#000000")

vuln_line_thousands
```

# Add number suffixes (Millions)

```{r}
vuln_line_millions <- ggplot(vuln_4_country, aes(x = year, y = vulnerable_pop, colour = country)) +
  geom_line() +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  labs(x = "Year", y = "Vulnerable Population (Millions)") +
  scale_y_continuous(labels = label_number(scale = 1e-6, suffix = "M")) +
  # Change the colours of lines to differentiate between countries
  scale_colour_manual(values = c("United Kingdom" = "#1E2A16",
                                 "Argentina" = "#D17E38",
                                 "Nepal" = "#444443",
                                 "Malawi" = "#3EA27F"),
  ) +
  # Remove legend set y-axis title to horizontal
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.06,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -150)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5)) +
  theme(legend.position = "none") +
  # Add labels to lines
  geom_text_repel(data = labels_df,
                  aes(label = country),
                  nudge_y = 500000,
                  colour = "#000000")

vuln_line_millions
```

# Add number suffixes (Billions)

```{r}
# Prepare the data
vuln_2_country <- vuln %>%
  # Select just the country, year and vulnerable_pop columns
  select(country, vulnerable_pop, year) %>%
  # India and China selected because there vulnerable populations are close to one billion
  filter(country %in% c("China", "India")) %>%
  group_by(country)

# Prepare data used to create labels
vuln_2_country_labels <- vuln_2_country %>%
  
  group_by(country) %>%
  
  slice_max(year)

vuln_line_billions <- ggplot(vuln_2_country, aes(x = year, y = vulnerable_pop, colour = country)) +
  geom_line() +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  labs(x = "Year", y = "Vulnerable Population (Billions)") +
  scale_colour_manual(values = c("China" = "#1E2A16",
                                 "India" = "#D17E38"
                                 ),
                      ) +
  # Remove legend set y-axis title to horizontal
  scale_y_continuous(labels = label_number(scale = 1e-9, suffix = "B")) +
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.06,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -150)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5)) +
  theme(legend.position = "none") +
  # Add labels to lines
  geom_text_repel(data = vuln_2_country_labels,
                  aes(label = country),
                  nudge_y = 500000,
                  colour = "#000000")

vuln_line_billions
```

## Focus chart

```{r}
vuln_focus <- vuln_4_country %>%
  mutate(highlight = country == country[which.max(vulnerable_pop)])

vuln_line <- ggplot(vuln_focus, aes(x = year, y = vulnerable_pop,  group = country, colour = highlight)) +
  geom_line(size = 1) +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  scale_colour_manual(values = c("TRUE" = "#48553F", "FALSE" = "#a6a6a6")) +
  scale_y_continuous(labels = label_number(scale = 1)) +
  # Remove legend set y-axis title to horizontal
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.06,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -70)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5)) +
  theme(legend.position = "none") +
  # Add labels to lines
  geom_text_repel(data = labels_df,
                  aes(label = country),
                  nudge_y = 500000,
                  colour = "#000000")

vuln_line
```

## Annotation

```{r}
vuln_line <- ggplot(vuln_4_country, aes(x = year, y = vulnerable_pop, colour = country)) +
  geom_line(size = 1) +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  scale_y_continuous(labels = label_number(scale = 1)) +
  # Change the colours of lines to differentiate between countries
  scale_colour_manual(values = c("United Kingdom" = "#1E2A16",
                                 "Argentina" = "#D17E38",
                                 "Nepal" = "#444443",
                                 "Malawi" = "#3EA27F"),
                      ) +
  # Remove legend set y-axis title to horizontal
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.06,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -70)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5)) +
  theme(legend.position = "none") +
  # Add labels to lines
  geom_text_repel(data = labels_df,
                  aes(label = country),
                  nudge_y = 500000,
                  colour = "#000000") +
  annotate("text", 
           x = 1997, 
           y = 1e7,
           label = "example text",
           colour = "#000000"
           )
  

vuln_line
```

## Break the start of the y-axis

```{r}
# If you are breaking the y-axis, make this clear to your users in your chart title, or in a note below your chart title
vuln_top_2 <- vuln %>%
  # Select just the country, year and vulnerable_pop columns
  select(country, vulnerable_pop, year) %>%
  # India and China selected so that a large section of the y-axis can be cut
  filter(country %in% c("China", "India")) %>%
  group_by(country)

  x_axis <- min(vuln_top_2$year)

# To Display the data

vuln_top_2

vuln_axis_break <- ggplot(vuln_top_2, aes(x = year, y = vulnerable_pop, colour = country)) +
  geom_line() +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  scale_colour_manual(values = c("China" = "#1E2A16",
                                 "India" = "#D17E38"
                                 )) +
  scale_y_continuous(limits = c(60000000,270000000),
                     breaks = seq(70000000,270000000, by = 40000000),
                     labels = label_number(scale = 1)
  ) + 
  coord_cartesian(clip = "off") +
  # Add break symbol
  annotate("segment", x = x_axis - 3, xend = x_axis - 1 , 
           y = 6.5e7, yend = 7e7, linewidth = 0.5) +
  annotate("segment", x = x_axis - 3, xend = x_axis - 1 , 
           y = 6e7, yend = 6.5e7, linewidth = 0.5) +
  # Drop vertical line down at x = 1950
  annotate(
    "segment",
    x = 1950, xend = 1950,
    y = -Inf, yend = 62500000,
    linewidth = 0.5
  ) +
  annotate(
    "segment",
    x = 1950, xend = 1950,
    y = 67500000, yend = Inf,
    linewidth = 0.5
  ) +
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.06,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -70)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5)) +
  theme(legend.position = "none") +
  # Add labels to lines
  geom_text_repel(data = vuln_2_country_labels,
                  aes(label = country),
                  nudge_y = 500000,
                  colour = "#000000")

vuln_axis_break
```

# x-axis break

```{r}
vuln_top_2_gap <- vuln %>%
  # Select just the country, year and vulnerable_pop columns
  select(country, vulnerable_pop, year) %>%
  filter(country %in% c("China", "India")) %>%
  group_by(country) %>%
  # Create artificial gap by removing vuln_pop for 1977
  mutate(vulnerable_pop = if_else(
    year %in% 1977,
    NA,
    vulnerable_pop
  )
  ) %>%
  ungroup()

vuln_top_2_gap

# Create data for labels

vuln_top_2_gap_labels <- vuln_top_2_gap %>%  
  
  group_by(country) %>%
  
  slice_max(year)
```

```{r}
vuln_line <- ggplot(vuln_top_2_gap, aes(x = year, y = vulnerable_pop, colour = country)) +
  geom_line() +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  annotate(
  "segment",
  x = 1972, xend = 1972,
  y = -Inf, yend = Inf,
  linewidth = 0.5,
  linetype = 2
) +
  annotate(
  "segment",
  x = 1982, xend = 1982,
  y = -Inf, yend = Inf,
  linewidth = 0.5,
  linetype = 2
) +
  annotate("text",
           x = 1977,
           y = 2.5e8,
           label = "data not available",
           colour = "#000000",
           size = 2.5) +
  annotate("text",
           x = 1977,
           y = 2.45e8,
           label = "for 1977",
           colour = "#000000",
           size = 2.5) +
  # Remove legend set y-axis title to horizontal
  scale_y_continuous(labels = label_number(scale = 1e-9, suffix = "B")) +
  scale_colour_manual(values = c("China" = "#1E2A16",
                                 "India" = "#D17E38"
                                 )) +
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.06,
                                    margin = margin(r = -75)
                                    ),
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5)) +
  theme(legend.position = "none") +
  # Add labels to lines
  geom_text_repel(data = vuln_top_2_gap_labels,
                  aes(label = country),
                  nudge_y = 500000,
                  colour = "#000000")

vuln_line
```

## Exporting (.svg file format)

```{r}
vuln_line <- ggplot(vuln_4_country, aes(x = year, y = vulnerable_pop, colour = country)) +
  geom_line() +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  scale_y_continuous(labels = label_number(scale = 1)) +
  # Change the colours of lines to differentiate between countries
  scale_colour_manual(values = c("United Kingdom" = "#1E2A16",
                                 "Argentina" = "#D17E38",
                                 "Nepal" = "#444443",
                                 "Malawi" = "#3EA27F"),
  ) +
  # Remove legend set y-axis title to horizontal
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                    vjust = 1.03,
                                    margin = margin(r = -80)
                                    ),
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5)) +
  theme(legend.position = "none") +
  # Add labels to lines
  geom_text_repel(data = labels_df,
                  aes(label = country),
                  nudge_y = 500000,
                  colour = "#000000",
                  segment.colour = NA)

# Note: the y-axis title will look incorrect in Rstudio but the saved .svg file will have the axis title in the correct position
# For the plot to look correct in Rstudio change vjust to 1.06
vuln_line

ggsave(
  # Set the file name
  filename = "vuln_line.svg",
  # Select the plot to save
  plot = vuln_line,
  # Set the save path
  path = "D:/Temp",
  # Set aspect ratio of plot, an aspect ratio of 16:9 will fill 1920x1080 displays
  width = 16,
  height = 9
)
```

# Exporting (.html file format) (currently has issues)

```{r}
#vuln_line <- ggplot(vuln_4_country, aes(x = year, y = vulnerable_pop, colour = country)) +
#  geom_line() +
#  scale_y_continuous(labels = label_number(scale = 1)) +
#  # Change the colours of lines to differentiate between countries
#  scale_colour_manual(values = c("United Kingdom" = "#1E2A16",
#                                 "Argentina" = "#D17E38",
#                                 "Nepal" = "#444443",
#                                 "Malawi" = "#3EA27F"),
# ) +
#  # Remove legend set y-axis title to horizontal
#  theme(axis.title.y = element_text(angle = 0, 
#                                    hjust = 1,
#                                    vjust = 1.03,
#                                    margin = margin(r = -80)
#                                    ),
#        plot.margin = margin(t = 20, l = 5, r = 5, b = 5)) +
#  theme(legend.position = "none")
#
## Note: the y-axis title will look incorrect in Rstudio but the saved file will have the axis title in the correct position
#vuln_line
#
#vuln_line_html <- ggplotly(vuln_line)
#
#saveWidget(vuln_line_html, file = "D:/Temp/vuln_line.html")
```

## Exercise

Create a line chart in R. Use the vulnerable dataset as your input data.

For an extra challenge, consider setting yourself an hour’s time limit to help simulate an emergency.

The line chart should:

- Show the size of the vulnerable populations of the 4 countries with the largest vulnerable populations in 2007 over the years from 1952 to 2007
-   Be formatted so that:
    -   The x-axis label is: Year 
    -   The y-axis label is: Vulnerable Population (Millions)
    -   The text size of the axis titles is 12
    -   The y-axis uses millions suffix (for example an axis break with a value of 6000000 would show as 6M)
    -   theme_minimal is used to set the background and gridline colours
    -   The vertical gridlines are removed
    -   The plot and panel backgrounds are HEX code #FFFFFF
    -   The line of the country with the largest population is HEX code #1E2A16
    -   The line of the country with the second largest population is HEX code #D17E38
    -   The line of the country with the third largest population is HEX code #444443
    -   The line of the country with the fourth largest population is HEX code #3EA27F
    -   Labels with country names are at the end of lines
    -   The chart is exported as an .svg file with the file name:
        vuln_line_solution
    -   The chart is saved to the D:/Temp folder
    -   The saved chart has an aspect ratio of 16:9

    
When you have finished making your line chart, compare it against the relevant code and solution in the 'exercise_solutions' subfolder. If you have followed all the guidance correctly, this is approximately how your output should look.