---
title: "Basic Data Visualisation, How to make bar charts in R: Exercise solution"
output: html_document
---

---
This block contains instructions on how to fill in this template.

**Please delete this block when creating your own file.**

This template should be used to create code snippet notebooks for each chapter
in a learning resource. Follow the instructions below to complete the template:
- Replace content in square brackets (e.g., [unit_name], [link], [chapter name])
with the relevant information. For example the title block below could be changed
from "[unit_name], [chapter_name] - Code snippets" to "Basic Data Visualisation,
Creating Bar Charts - Code snippets".
- In the setup section add any instructions for users to follow before running the
code snippets, such as installing packages or downloading data files. Add any
necessary import statements or setup code in the first code cell.
- Code snippets should be grouped into sections according to the sections in the
learning resource.
- Each code snippet from the learning resource should be placed in its own code
cell under the relevant section, with any necessary context provided in a
markdown block above each code cell.
- Follow [tidyverse](https://style.tidyverse.org/) style guide for R code
formatting.
- Whenever you need to reference a file in [learning_resources](http://_vscodecontentref_/1), use:
```r
file_path <- file.path(learning_resources_dir, "data", "filename.csv")
```

- Definitions:
    - "Unit name" refers to the unit name, which is also the name of the folder
    this python folder is in.
    - "Chapter name" refers to the specific chapter that the code snippets are
    in. Chapters are the highest level headings in the learning resource.
    - "Section name" refers to subheadings within chapters that code snippets
    are in.
---

# Basic Data Visualisation, How to make bar charts in R: Exercise solution

The following code relates to the **Making bar charts in R** script in the
**Basic Data Visualisation** learning resource on the [Analysis for Action]([PLACEHOLDER])
platform.

This is the code used to produce the exercise solution from the **Making bar charts in R** script. Compare your code and output against the code and output produced here to see how well you did. Note that your code and output do not need to match the solution exactly to be considered correct!

Can you think of any other ways of making the bar chart more accessible?

The bar chart should:

-   Show the size of vulnerable populations in Argentina, Malawi, Nepal, and the United Kingdom in the year 1997
-   Be formatted so that:
    -   The x-axis label is: Country 
    -   The y-axis label is: Vulnerable Population (Millions) 
    -   The text size of the axis titles is 12
    -   The panel background and plot background is HEX code #FFFFFF
    -   The largest bar is HEX code #324028, the rest are HEX code #D4DDD6
    -   The width of the bars is 0.6
    -   The bars are ordered from largest to smallest
    -   The horizontal gridlines are HEX code #D9D9D9 and size = 0.5
    -   There are no vertical gridlines
    -   The y-axis uses millions suffix (for example an axis break with a value of 6000000 would show as 6M)
    -   The chart is exported as an .svg file with the file name:
        vuln_pop_partner_countries 
    -   The chart is saved to the D:/Temp folder
    -   The save charted has an aspect ratio of 16:9

## Troubleshooting

If you encounter any issues running the code snippets, consider the following
troubleshooting steps:
- Ensure you have followed the setup instructions in the
[README.md](../../../README.md) file including `install.packages()` for
required packages.
- Run code from top to bottom to ensure all dependencies are met.
- Restart your R session and clear all outputs.
- For more detailed troubleshooting guidance, refer to the
[troubleshooting document]([PLACEHOLDER]).

## Setup

Before running the code snippets, ensure you have:
- followed the setup instructions in the [README.md](../../../README.md) file,
- **[add further instructions as needed]**,
- run the code block below.

```{r setup, message=FALSE}
# This chunk should include:
# * library() calls for required packages
# * paths to data files
# * any other setup required

library(readr)
library(here)
library(dplyr)
library(tidyr)
library(janitor)
library(gt)
library(ggplot2) # The scripts uses ggplot2 version 4.0.0
library(svglite)
library(scales)
library(plotly)
library(htmlwidgets)

# Use here::here() to construct paths, do not setwd()
here::here()
vuln <- read.csv(here::here("learning_resources", "data", "vulnerable.csv"))
```

## Preparing the dataframe

```{r}
# Prepare your data to make it suitable for a bar chart

vuln_1997_partner_countries <- vuln %>%
  
  filter(year %in% ("1997")) %>%
  
  filter(country %in% c("Argentina", "Malawi", "Nepal", "United Kingdom")) %>%
  
  select(country, vulnerable_pop) 

# Order the data by population from largest to smallest

vuln_1997_partner_countries <- vuln_1997_partner_countries %>% 
  dplyr::arrange(desc(vulnerable_pop))

vuln_1997_partner_countries$country <- factor(
  vuln_1997_partner_countries$country,
  levels = vuln_1997_partner_countries$country
)

# Create the data for the focus chart, extra row added which has TRUE for the largest population and FALSE for the others

vuln_focus_1997 <- vuln_1997_partner_countries %>%
  mutate(highlight = vulnerable_pop == max(vulnerable_pop))
```

# Creating the plot

```{r}
# Create the plot and specify the data on the x and y axis and set the fill for the focus chart
vuln_bar <- ggplot(vuln_focus_1997, aes(x = country, y = vulnerable_pop, fill = highlight)) +
  # Set bar width
  geom_bar(stat = "identity", 
           width = 0.6) +
  # Use theme_minimal to increase contrast
  theme_minimal() +
  # Remove vertical gridlines
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  # Add number suffixes
  scale_y_continuous(labels = label_number(scale = 1e-6, suffix = "M")) +
  # Set the title and axis titles
  labs(
    x = "Country", 
    y = "Vulnerable Population (Millions)") +
  # Set the text sizes
  theme(
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
  ) +
  # Set the colours for the focus chart
  scale_fill_manual(values = c("TRUE" = "#324027", "FALSE" = "#949494"),
                    guide = "none") +
  # Set y-axis title to horizontal
  theme(axis.title.y = element_text(angle = 0, 
                                    hjust = 1, 
                                    vjust = 1.03,
                                    # Adjust margin to move the title to the right
                                    margin = margin(r = -150)
                                    ),
        # Adjust plot margin
        plot.margin = margin(t = 20, l = 5, r = 5, b = 5))
```

# Save as .html (has issues)

```{r}
#vuln_bar_html <- ggplotly(vuln_bar)
#
#saveWidget(vuln_bar_html, file = "D:/Temp/vuln_pop_partner_countries.html")
```

# Save as .svg

```{r}
ggsave(
  # Set the file name
  filename = "vuln_pop_partner_countries.svg",
  # Select the plot to save
  plot = vuln_bar,
  # Set the save path
  path = "D:/Temp",
  # Set aspect ratio of plot, an aspect ratio of 16:9 will fill 1920x1080 displays
  width = 16,
  height = 9
)
```

